pyTopoComplexity: A Python package for topographic complexity analysis

### 2024-08-29 JOSS/paper.rmd
Add files via upload

**pyTopoComplexity** is a Python package that provides a computationally efficient and customizable 
implementation of four modern methods for quantifying topographic complexity. These methods include 
two-dimensional continuous wavelet transform (2D-CWT) analysis, fractal dimension estimation, Rugosity 
Index and Terrain Position Index calculations across various spatial scales. This package addresses 
the scarcity of open-source software for these sophisticated methods needed in modern 
terrain analysis, and facilitates data comparison and reproducibility. In the [software respository](https://github.com/GeoLarryLai/pyTopoComplexity.git), 
we also include a Jupyter Notebook file that integrates components from the Python-based 
surface-process modeling platform **Landlab** [@Hobley2017]. The notebook allows researchers to 
simulate the smoothing of topography over time through terrestrial nonlinear hillslope 
diffusion processes. By combining these features, **pyTopoComplexity** advances the toolset 
available to researchers for measuring and simulating the time-dependent persistence of 
topographic complexity signatures against environmental forces on terrain surfaces.

@@

Topographic complexity, often referred to as topographic roughness or surface roughness, provides 
critical insights into surface processes and the interactions among the geosphere, biosphere, and 
hydrosphere [@Dietrich2006]. With the increasing availability, utility, and popularity of digital terrain model (DTM) 
data, quantifying topographic complexity has become an essential step in terrain analysis across 
various research fields. This necessity spans applications such as terrain classification and mapping 
at various spatial scales [@Weiss2001; @Robbins2018; @Lindsay2019; @PardoIguzquiza2022a], evaluating 
the depositional age of event sedimentation (e.g., landslides, alluvial fans) and subsequent erosion processes 
[@Hetz2016; @Johnstone2018; @Booth2017; @LaHusen2020; @Herzig2023], and identifying habitats to assess 
ecological diversity on land and seafloor [@Frost2005; @Hetz2016; @Wilson2007].

In recent years, several advanced methods for quantifying topographic complexity have been developed, 
including two-dimensional continuous wavelet transform (2D-CWT) analysis [@Booth2009; @Berti2013], 
fractal dimension estimation [@Taud2005; @Robbins2018; @PardoIguzquiza2020], and Rugosity Index 
calculation [@Jenness2004; @DuPreez2015]. These methods are considered more effective for terrain analysis 
tasks compared to conventional morphological metrics such as variations in local slope and relief. 
Despite their importance, comprehensive publicly available tools that incorporate these advanced 
methods for studying topographic complexity are lacking. Common open-source geospatial analysis software, 
such as QGIS [@QGIS_software], GRASS GIS [@GRASS_GIS_software], and WhiteboxTools [@Lindsay2016], only 
implement basic conventional methods, limiting the reproducibility and comparability of these newer 
approaches. Although some specialized programs for calculating the Rugosity Index exist [@Walbridge2018; @Benham2022], 
they have been limited to marine bathymetric studies and involve various mathematical choices and designs.

To address this gap, we have developed an open-source Python toolkit called **pyTopoComplexity**. This toolkit 
offers computationally efficient and easily customizable implementations for performing and visualizing the results 
of 2D-CWT, fractal dimension, and rugosity index calculations (see Table 1\autoref{tab:1}). Additionally, **pyTopoComplexity** 
includes a module for calculating the Terrain Position Index (TPI), a widely used metric in geomorphology 
research [@Newman2018; @Deumlich2010; @Liu2011] and often used alongside Rugosity Index in marine 
geological and ecological studies [@Wilson2007; @Walbridge2018].

@@

The **pycwtmexhat.py** module in **pyTopoComplexity** implements the 2D-CWT method for terrain analysis, providing detailed information on how 
amplitude is distributed across spatial frequencies at each position in the data by transforming spatial data into position-frequency 
space. When used with the Gaussian family of wavelets, this method is particularly effective for depicting the Laplacian of topography 
[@Torrence1998; @Lashermes2007], revealing concave and convex regions of topography at various smoothing-length scales [@Malamud2001; @Struble2021], 
identifying deep-seated landslides [@Booth2009; @Berti2013], and estimating the depositional ages of landslide deposits [@Booth2017; @LaHusen2020; @Underwood2022; @Herzig2023]. 
Unlike techniques that use moving windows to detect regional topographic complexity, the 2D-CWT method stands out by isolating specific landform features at the scale of the 
designated $\lambda$ while filtering out noise from terrain variations at longer or shorter wavelengths.

@@

, where the resultant wavelet coefficient $C(s,x,y)$ provides a measure of how well the wavelet $\psi$ matches the data $z$ at each grid point [@Torrence1998]. 
When $s$ is large, $\psi$ is spread out, capturing long-wavelength features of $z$; when $s$ is small, $\psi$ becomes more localized, making it sensitive 
to fine-scale features of $z$. In this implementation, we use the 2D Ricker or Marr wavelet (i.e., Mexican hat wavelet) function to define $\psi$:

@@

The Mexican hat wavelet as defined above is equal to the second derivative of a Gaussian envelope, with its Fourier wavelength ($\lambda$) 
dependent on the chosen wavelet scale ($s$) and the grid spacing ($\Delta$) of the input DTM raster. The wavelet function $\psi$ is scaled according to the 
wavelet scale parameter $s$ and the grid spacing $\Delta$, ensuring that the resultant wavelet coefficient $C$ signifies 
concave and convex landforms corresponding to the wavelet scale $s$. Users can define the value of $\lambda$ in meters as the 
targeted spatial scale for landform roughness analysis, and the **pycwtmexhat.py** module will automatically compute the wavelet 
scale $s$ based on the grid spacing ($\Delta$) of the input raster file (Figure 1)\autoref{fig:1}.

We note that the equations for $C$ and $\psi$ presented here are mathematical approaches adapted in later publications 
[@LaHusen2020; @Underwood2022; @Herzig2023] and ongoing works on landslide mapping and age dating studies [@Booth2023; @Lai2023; @Ozioko2023]. 
There are minor differences in the proportionality constant used to define $\psi$ and the conventions used to present the magnitude of the wavelet 
coefficients $C$, compared to earlier similar research by @Booth2009 and @Booth2017 (original MATLAB codes 
available from [Booth's personal website](https://web.pdx.edu/~boothad/tools.html)). These differences in methatical approach will by definition 
result in $C$ values in different units and order of magnitude (e.g., $10^{-3}$ to $10^{-4}$ [m$^{-2}$] in @Booth2017 
and prior studies; $10^{-2}$ to $10^{-3}$ [m$^{-1}$] in @LaHusen2020 thereafter). Despite this discrepancy, the $C$ values 
yielded from these two approaches are linearly scaled and interconvertible, and they both reflect identical spatiotemporal patterns 
of topographic complexity (i.e., surface roughness).

![Figure 1. Hillshade map of the 2014 Oso Landslide region (upper-left) and the results of the two-dimensional continuous wavelet 
transform (2D-CWT) analysis using the **pycwtmexhat.py** module, with designated Fourier wavelengths ($\lambda$) of the Mexican Hat 
wavelet at 15 m, 30 m, 45 m, 60 m, and 75 m. \label{fig:1}](fig1-pycwtmexhat.png){ width=100% }

@@

, where $Z(i)$ is the elevation at location $i$ along the profile. The local fractal dimension ($FD$) is estimated from one-dimensional 
profiles in principal directions (i.e., horizontal, vertical, and diagonal) within a square moving window. Assuming that fractional 
Brownian motion is an appropriate stochastic model for natural surfaces, its variogram follows a power-law model with respect to $k$ [@Wen1997]:

@@

, and its exponent $\beta$ is related to the local fractal dimension ($FD$) by:

$$
FD = TD + 1 - \frac{\beta}{2}
$$

, where $TD$ is the topological dimension in the Euclidean space of the fractional Brownian motion. For one-dimensional fractional Brownian 
motion, $TD = 1$. The fractal dimension of the two-dimensional surface $(FD)_2^*$ can be estimated as the average fractal dimension of 
the one-dimensional profiles $(FD)_1^*$:

$$
(FD)_2^* = 1 + (FD)_1^*
$$

@@

$$
\text{Rugosity Index} = \frac{\text{contoured area}}{\text{planimetric area}}
$$

@@

By definition, the Rugosity Index is as a minimum value of one (completely flat surface). Typical values of the conventional rugosity 
index (without slope correction) range from one to three although larger values are possible in very steep terrains. The slope-corrected rugosity 
index, also called arc-chord ratio (ACR) Rugosity Index, could provide a better representation of local surface complexity because it is not biased by slope (Figure 3)\autoref{fig:3}.

![Figure 3. Calculated results of the Arc-Chord Ratio Rugosity Index (upper) and the conventional Rugosity Index (lower) for the 2014 Oso Landslide 
region, using the **pyrugosity.py** module. The calculations were performed with different moving window sizes: 17 by 17 grids (left), 49 by 49 grids 
(middle), and 83 by 83 grids (right). Grid spacing is 3 U.S. survey feet (approximately 0.9144 meters). \label{fig:3}](fig3-pyrugosity.png){ width=100% }

@@

$$
TPI = Z_{\text{grid}} - \text{mean}(Z_{\text{neighborhood}})
$$

In this module, the TPI is calculated for the central grid within a square moving window. Users can specify the size of the window (i.e., the 
number of grids along each edge) to evaluate topographic positions at various spatial scales. Positive TPI values indicate generally convex, 
elevated features, while negative values represent concave depressions. Values close to zero denote relatively planar areas. The **pytpi.py** 
module also returns the absolute values of the TPI (Figure 4\autoref{fig:4}). This absolute TPI raster is suggested to be effective for 
quick assessments in mapping and relative dating of deep-seated landslide deposits using high-resolution DTM data [@Lai2023].

![Figure 4. Calculated results of the Terrain Position Index (upper) and its absolute values (lower) for the 2014 Oso Landslide region, 
using the **pytpi.py** module. The calculations were performed with different moving window sizes: 17 by 17 grids (left), 49 by 49 grids 
(middle), and 83 by 83 grids (right). Grid spacing is 3 U.S. survey feet (approximately 0.9144 meters). \label{fig:4}](fig4-pytpi.png){ width=100% }

>>>>

**pyTopoComplexity** is a Python package that provides a computationally efficient and customizable 
implementation of four modern methods for quantifying topographic complexity. These methods include 
two-dimensional continuous wavelet transform analysis, fractal dimension estimation, Rugosity 
Index and Terrain Position Index calculations across various spatial scales. This package addresses 
the scarcity of open-source software for these sophisticated methods needed in modern 
terrain analysis, and facilitates data comparison and reproducibility. In the [software respository](https://github.com/GeoLarryLai/pyTopoComplexity.git), 
we also include a Jupyter Notebook file that integrates components from the Python-based 
surface-process modeling platform **Landlab** [@Hobley2017]. The notebook allows researchers to 
simulate the smoothing of topography over time through terrestrial nonlinear hillslope 
diffusion processes. By combining these features, **pyTopoComplexity** advances the toolset 
available to researchers for measuring and simulating the time-dependent persistence of 
topographic complexity signatures against environmental forces on terrain surfaces.

@@

Topographic complexity, often referred to as topographic roughness or surface roughness, provides 
critical insights into surface processes and the interactions among the geosphere, biosphere, and 
hydrosphere [@Dietrich2006]. With the increasing availability, utility, and popularity of digital terrain model (DTM) 
data, quantifying topographic complexity has become an essential step in terrain analysis across 
various research fields. This necessity spans applications such as terrain classification and mapping 
at various spatial scales [@Weiss2001; @Robbins2018; @Lindsay2019; @PardoIguzquiza2022a], evaluating 
the depositional age of event sedimentation (e.g., landslides, avulsion on alluvial fans) and subsequent erosion processes 
[@Hetz2016; @Johnstone2018; @Booth2017; @LaHusen2020; @Herzig2023], and identifying habitats to assess 
ecological diversity on land and seafloor [@Frost2005; @Hetz2016; @Wilson2007].

In recent years, several advanced methods for quantifying topographic complexity have been developed, 
including two-dimensional continuous wavelet transform (2D-CWT) analysis [@Booth2009; @Berti2013], 
fractal dimension estimation [@Taud2005; @Glenn2006; @Robbins2018; @PardoIguzquiza2020], and Rugosity Index 
calculation [@Jenness2004; @DuPreez2015]. These methods are considered more effective for topographic 
complexity analysis tasks compared to conventional morphological metrics such as variations in local slope and relief. 
Despite their importance, comprehensive publicly available tools that incorporate these advanced 
methods for studying topographic complexity are lacking. Common open-source geospatial analysis software, 
such as QGIS [@QGIS_software], GRASS GIS [@GRASS_GIS_software], and WhiteboxTools [@Lindsay2016], only 
implement basic conventional methods, limiting the reproducibility and comparability of these newer 
approaches. Although specialized programs for calculating the Rugosity Index exist [@Walbridge2018; @Benham2022], 
they have been confined to marine bathymetric studies and involve various mathematical limitations, assumptions, and designs.

To address this gap, we have developed an open-source Python toolkit called **pyTopoComplexity**. This toolkit 
offers computationally efficient and easily customizable implementations for performing and visualizing the results 
of 2D-CWT, fractal dimension, and Rugosity Index calculations (see Table 1\autoref{tab:1}). Additionally, **pyTopoComplexity** 
includes a module for calculating the Terrain Position Index, a widely used metric in geomorphology 
research [@Newman2018; @Deumlich2010; @Liu2011] and often used alongside Rugosity Index in marine 
geological and ecological studies [@Wilson2007; @Walbridge2018].

@@

@@

, where the resultant wavelet coefficient $C(s,x,y)$ provides a measure of how well the wavelet $\psi$ matches the data $z$ at each grid point [@Torrence1998]. 
Unlike other techniques that use moving windows to detect regional topographic complexity, the 2D-CWT method stands out by isolating specific landform features 
at the scale of the designated wavelength while filtering out noise from terrain variations at longer or shorter wavelengths. When $s$ is large, $\psi$ is spread out, capturing long-wavelength features of $z$; when $s$ is small, $\psi$ becomes more localized, making it sensitive 
to fine-scale features of $z$. In this implementation, we use the 2D Ricker or Marr wavelet (i.e., Mexican hat wavelet) function to define $\psi$ [@Ricker1943]:

@@

The Mexican hat wavelet, as defined above, corresponds to the second derivative of a Gaussian envelope. Its Fourier wavelength ($\lambda$) 
depends on the chosen wavelet scale ($s$) and the grid spacing ($\Delta$) of the input DTM raster. The wavelet function $\psi$ is scaled based on the wavelet 
scale parameter $s$ and the grid spacing $\Delta$, ensuring that the resultant wavelet coefficient $C$ represents concave and convex landforms according to the 
wavelet scale $s$. Users can specify the wavelength $\lambda$ in meters as the target spatial scale for landform roughness analysis. The **pycwtmexhat.py** 
module will automatically compute the wavelet scale $s$ based on the grid spacing ($\Delta$) of the input raster file (Figure 1\autoref{fig:1}). In this module, 
users can choose to perform convolution either in the original domain (i.e., direct convolution) or in the frequency domain (i.e., using the Fast Fourier Transform) 
with the `convolve2d` and `fftconvolve` functions from the **SciPy** package, respectively [@2020SciPy-NMeth]. By default, the module uses `fftconvolve` for greater 
computational efficiency.

We note that the equations for $C$ and $\psi$ presented here are mathematical approaches adapted in later publications 
[@LaHusen2020; @Underwood2022; @Herzig2023] and ongoing works on landslide mapping and age dating studies [@Booth2023; @Lai2023; @Ozioko2023]. 
There are minor differences in the proportionality constant used to define $\psi$ and the conventions used to present the magnitude of the wavelet 
coefficients $C$, compared to earlier similar research by @Booth2009 and @Booth2017 (original MATLAB codes 
available from [Booth's personal website](https://web.pdx.edu/~boothad/tools.html)). These differences in mathematical approach will by definition 
result in $C$ values in different units and order of magnitude (e.g., $10^{-3}$ to $10^{-4}$ [m$^{-2}$] in @Booth2017 
and prior studies; $10^{-2}$ to $10^{-3}$ [m$^{-1}$] in @LaHusen2020 thereafter). Despite this discrepancy, the $C$ values 
yielded from these two approaches are linearly scaled and interconvertible, and they both reflect identical spatiotemporal patterns 
of topographic complexity (i.e., surface roughness).

![Figure 1. Hillshade map of the 2014 Oso Landslide region (upper-left) and the results of the two-dimensional continuous wavelet 
transform (2D-CWT) analysis using the **pycwtmexhat.py** module, with designated Fourier wavelengths ($\lambda$) of the Mexican Hat 
wavelet at 15 m, 45 m, and 75 m. \label{fig:1}](fig1-pycwtmexhat.png){ width=100% }

@@

, where $Z(i)$ is the elevation at location $i$ along the profile. The local fractal dimension ($\text{FD}$) is estimated from one-dimensional 
profiles in principal directions (i.e., horizontal, vertical, and diagonal) within a square moving window. Assuming that fractional 
Brownian motion is an appropriate stochastic model for natural surfaces, its variogram follows a power-law model with respect to $k$ [@Wen1997]:

@@

, and its exponent $\beta$ is related to the local fractal dimension ($\text{FD}$) by:

$$
\text{FD} = \text{TD} + 1 - \frac{\beta}{2}
$$

, where $\text{TD}$ is the topological dimension in the Euclidean space of the fractional Brownian motion. For one-dimensional fractional Brownian 
motion, $\text{TD} = 1$. The fractal dimension of the two-dimensional surface $(\text{FD})_2^*$ can be estimated as the average fractal dimension of 
the one-dimensional profiles $(\text{FD})_1^*$:

$$
(\text{FD})_2^* = 1 + (\text{FD})_1^*
$$

@@

$$
\text{Rugosity Index} = \frac{\text{Contoured area}}{\text{Planimetric area}}
$$

@@

By definition, the Rugosity Index has a minimum value of one (completely flat surface). Typical values of the conventional rugosity 
index (without slope correction) range from one to three although larger values are possible in very steep terrains. The slope-corrected rugosity 
index, also called arc-chord ratio (ACR) Rugosity Index, could provide a better representation of local surface complexity because it is not biased by slope (Figure 3)\autoref{fig:3}.

![Figure 3. Calculated results of the Arc-Chord Ratio Rugosity Index (left) and the conventional Rugosity Index (right) for the 2014 Oso Landslide 
region, using the **pyrugosity.py** module. The calculations were performed with different moving window sizes: 17 by 17 grids (top) and 49 by 49 grids 
(bottom). Grid spacing is 3 U.S. survey feet (approximately 0.9144 meters). \label{fig:3}](fig3-pyrugosity.png){ width=100% }

@@

$$
\text{TPI} = Z_{\text{grid}} - \text{mean}(Z_{\text{neighborhood}})
$$

In this module, the TPI is calculated for the central grid within a square moving window. Users can specify the size of the window (i.e., the 
number of grids along each edge) to evaluate topographic positions at various spatial scales. Positive TPI values indicate generally convex, 
elevated features, while negative values represent concave depressions (i.e., saddle features). Values close to zero denote relatively flat surface 
or area with near continuous slope. The **pytpi.py** module also returns the absolute values of the TPI (Figure 4\autoref{fig:4}). This absolute TPI calculation is suggested to be effective for 
quick assessments in mapping and relative dating of deep-seated landslide deposits using high-resolution DTM data [@Lai2023].

![Figure 4. Calculated results of the Terrain Position Index (left) and its absolute values (right) for the 2014 Oso Landslide region, 
using the **pytpi.py** module. The calculations were performed with different moving window sizes: 17 by 17 grids (top) and 83 by 83 grids (bottom). 
Grid spacing is 3 U.S. survey feet (approximately 0.9144 meters). \label{fig:4}](fig4-pytpi.png){ width=100% }
